\large{\bf Looking forwards and backwards through locally regulated populations }

### 2023-03-22 forwards-and-backwards.tex
reinserted old lemma 6.1

If $r(x,m)$ is independent of $m$, then the conditions are easy to 
satisfy; they just require some regularity of $r$ as a function of $x$. 
Condition~\ref{control through r} is also satisfied if for example
$\|\nabla\rho_r\|\leq C\rho_r$ 
and $m\partial_mr(x,m)$, $m^2\partial_{mm}r(x,m)$ are bounded.
This is the case, for instance, if $\rho_r$ decays exponentially.
On the other hand, it might seem more natural to take $\rho_r$ to be a Gaussian 
density with parameter $\sigma_r$, say. Then, 
as we check in Lemma~\ref{old 6.1}, \comment{this is referring to the
old Lemma 6.1, now to go in Appendix,}
Condition~\ref{control through gamma} is satisfied if $\rho_\gamma$ is also
Gaussian with parameter $\sigma_\gamma$ and $\sigma_\gamma>\sigma_r$. For large enough
$\theta$, this condition guarantees that 
$\sigma_r+1/\theta <\sigma_\gamma$, so that the establishment probability of a juvenile
is controlled by individuals that are already `felt' by the fecundity kernel $\rho_\gamma$
at the location of their parent.

@@

>>>>

If $r(x,m)$ is independent of $m$, then the conditions are easy to 
satisfy; they just require some regularity of $r$ as a function of $x$. 
Condition~\ref{control through r} is also satisfied if for example
$\|\nabla\rho_r\|\leq C\rho_r$ 
and $m\partial_mr(x,m)$, $m^2\partial_{mm}r(x,m)$ are bounded.
This is the case, for instance, if $\rho_r$ decays exponentially.
On the other hand, it might seem more natural to take $\rho_r$ to be a Gaussian 
density with parameter $\sigma_r$, say. Then, 
as we check in Lemma~\ref{lem:gamma_bound},
Condition~\ref{control through gamma} is satisfied if $\rho_\gamma$ is also
Gaussian with parameter $\sigma_\gamma$ and $\sigma_\gamma>\sigma_r$. For large enough
$\theta$, this condition guarantees that 
$\sigma_r+1/\theta <\sigma_\gamma$, so that the establishment probability of a juvenile
is controlled by individuals that are already `felt' by the fecundity kernel $\rho_\gamma$
at the location of their parent.

@@

\subsection{Constraints on kernel widths}

\begin{lemma}
    \label{lem:gamma_bound}
\comment{TODO; REWRITE TO SAY THAT STHG IMPLIES CONDITION 2 OF LEMMA \ref{lem:conditions on r} (see reference on p.13)}
    Under the assumptions of Definition~\ref{def:model_setup},
    for all $f \in \mathcal{C}^{2}_{0}(\IR^d \times [0,\infty))$,
    there exists a $K_f < \infty$ such that
    for all $\theta > 0$ and all $\eta \in \measures$,
    \begin{equation} \label{eqn:gamma_bound}
    \gamma\big(x,\smooth{\gamma} \eta(x)\big)
    \left| \theta 
        \int_{\IR^n} 
            \left\{
                f(y, \smooth{r} \eta(y)) - f(x, \smooth{r} \eta(x))
            \right\}
        q_{\theta}(x,dy)
    \right|
    \leq C_f     .
    \end{equation}
\end{lemma}

The key assumption here from Definition~\ref{def:model_setup}
will be that the spatial scale over which local density affects birth rate
is larger than the scale over which it affects establishment:
concretely, that $\epsilon_{r}^2 + \frac{2\lambda}{\theta} < \epsilon_{\gamma}^2$,
where $\lambda = \sup_x \sup_{y : \|y\| = 1} y^T \covq(x) y$
is the largest eigenvalue of $\covq(x)$ across all $x$.
In the simple case of $\meanq = 0$ and $\covq = \sigma^2 I$,
the condition is simply that $\epsilon_r^2 + 2\sigma^2 / \theta < \epsilon_\gamma^2$.

\begin{proof}[of Lemma \ref{lem:gamma_bound}]

    Since we have assumed that there is a $K$ for which $m^2 \gamma(x, m) < K$
    for all $x \in \IR^d$ and $m \ge 0$,
    equation~\eqref{eqn:gamma_bound} will follow if we can show that
    \begin{equation*}
        \left| \theta 
            \int_{\IR^n} 
                \left\{
                    f\left(y, \smooth{r}\eta(y)\right)
                    -
                    f\left(x, \smooth{r}\eta(x)\right)
                \right\}
            q_{\theta}(x,dy)
        \right|
        \leq
        C_f \left( 1 + \smooth{\gamma} \eta(x)^2 \right) .
    \end{equation*}
    To do this, first note that 
    \begin{align} \label{eqn:q_bound_split}
    &
        \left| \theta 
            \int_{\IR^n} 
                \left\{
                    f\left(y, \smooth{r}\eta(y)\right)
                    -
                    f\left(x, \smooth{r}\eta(x)\right)
                \right\}
            q_{\theta}(x,dy)
        \right|
    \\ &\qquad {} \leq
        \left| \theta 
            \int_{\IR^n} 
                \left\{
                    f\left(y, \smooth{r}\eta(y)\right)
                    -
                    f\left(x, \smooth{r}\eta(y)\right)
                \right\}
            q_{\theta}(x,dy)
        \right|
    \\ &\qquad \qquad {}  +
        \left| \theta 
            \int_{\IR^n} 
                \left\{
                    f\left(x, \smooth{r}\eta(y)\right)
                    -
                    f\left(x, \smooth{r}\eta(x)\right)
                \right\}
            q_{\theta}(x,dy) 
        \right| .
    \end{align}

    Since 
    \begin{align*}
        f\left(y, \smooth{r}\eta(y)\right) - f\left(x, \smooth{r}\eta(y)\right)
        &=
        \sum_i (y_i - x_i) \partial_{x_i} f \left(x, \smooth{r}\eta(y)\right) 
        \\ &\qquad {}
        +
        \frac{1}{2} \sum_{ij} (y_i - x_i)(y_j - x_j)
            \partial_{x_i x_j} f \left(z, \smooth{r}\eta(y)\right) 
    \end{align*}
    where $z = t x + (1-t) y$ for some $0 \le t \le 1$,
    writing
    $K_1 = \sup_{x,m} \max_i |\partial_{x_i} f(x, m)|$
    and 
    $K_2 = \sup_{x,m} \max_{i,j} |\partial_{x_i x_j} f(x, m)|$
    for the bounds on the spatial derivatives of $f$,
    the first term is bounded by
    \begin{align*}
        K_1 \left|
            \theta \int_{\IR^d} \sum_i (y_i - x_i) q_\theta(x, dy)
        \right|
        +
        K_2 \left|
            \theta \int_{\IR^d} \sum_{ij} (y_i - x_i) (y_j - x_j)  q_\theta(x, dy)
        \right| .
    \end{align*}
    This is bounded uniformly because
    $q_\theta(x, dy)$ is the density of a Gaussian with mean $\meanq(x)/\theta$
    and covariance $\covq(x)/\theta$, and both $\meanq(x)$ and $\covq(x)$
    are bounded uniformly in $x$.

    Similarly, 
    \begin{align*}
        f(x, \smooth{r}\eta(y)) - f(x, \smooth{r}\eta(x))
        &=
        (\smooth{r}\eta(y) - \smooth{r}\eta(x)) f'(x, \smooth{r}\eta(x))
        + \frac{1}{2} (\smooth{r}\eta(y) - \smooth{r}\eta(x))^2 f''(x, m)
    \end{align*}
    for some $\smooth{r}\eta(x) \le m \le \smooth{r}\eta(y)$,
    where $f'$ and $f''$ denote the first and second derivatives with respect to the second argument.
    So,
    the second term in~\eqref{eqn:q_bound_split} is bounded by
    \begin{align*}
        % note that the f' term does not depend on x, just y,
        % so can come out of the integral
        \|f'\|_\infty
        \left|
            \theta \int_{\IR^d}
                (\smooth{r}\eta(y) - \smooth{r}\eta(x))
            q_\theta(x, dy)
        \right|
        +
        \|f''\|_\infty
        \left|
            \theta \int_{\IR^d}
                (\smooth{r}\eta(y) - \smooth{r}\eta(x))^2
            q_\theta(x, dy)
        \right| .
    \end{align*}
    Therefore, it suffices to show that there is a $K < \infty$ such that
    for all $\eta \in \measures$ and $x \in \IR^d$,
    \begin{equation}
        \label{eqn:first_moment_rho}
        \left|\theta \int_{\IR^d}
            ( \smooth{r} \eta(y)-\smooth{r} \eta(x) )
        q_\theta(x,dy)\right|
        \leq K \smooth{\gamma} \eta(x)
    \end{equation}
    and
    \begin{equation}
        \label{eqn:second_moment_rho}
        \theta \int_{\IR^d}
                \left( \smooth{r} \eta(y) - \smooth{r} \eta(x) \right)^2
        q_\theta(x,dy)
        \leq K \left(\smooth{\gamma}\eta(x) \right)^2 .
    \end{equation}
    Note that the right hand side of each is the average density
    over a wider region (since $\epsilon_\gamma > \epsilon_r$).

    First we prove \eqref{eqn:first_moment_rho}.
    Recall that $\smooth{r} \eta(x) = \int p_{\epsilon^2_r}(x - w) \eta(dw)$,
    where $p_t$ is the density of a Gaussian with mean 0 and variance $t$,
    so that applying Fubini, \eqref{eqn:first_moment_rho} is
    \begin{align*}
        \left|
        \int_{\IR^d}
        \theta
            \int_{\IR^d}
                ( p_{\epsilon_r^2}(y - w) - p_{\epsilon_r^2}(x - w) )
            q_\theta(x, dy)
        \eta(dw)
        \right| .
    \end{align*}
    Write $p_{s, x}(\cdot)$ for the density of a Gaussian
    with mean $s \meanq(x)$ and covariance $\epsilon_r^2 I + s \covq(x)$,
    so that $\int p_{\epsilon_r^2}(y - w) q_\theta(x, dy) = p_{1/\theta, x}(w-x)$.
    It therefore suffices to show that for all $x$ and $w \in \IR^d$,
    there exists $K$ such that
    \begin{align} \label{eqn:goal1}
        \left|
        \theta
            \int_{\IR^d}
                ( p_{\epsilon_r^2}(y-w) - p_{\epsilon_r^2}(x-w) )
            q_\theta(x, dy)
        \right|
        &=
        \theta \left|
            p_{1/\theta, x}(w-x) - p_{0, x}(w-x)
        \right|
        \le
        K p_{\epsilon^2_\gamma}(w-x) .
    \end{align}
    However,
    $\theta( p_{1/\theta, x}(z) - p_{0, x}(z) )
    = \partial_s p_{s, x}(z)$ for some $0 \le s \le 1/\theta$.
    Write $\Gamma(s,x) = \epsilon_r^2 I + s \covq(x)$,
    so that
    \begin{align*}
        p_{s, x}(z)
        &=
        \frac{1}{\left(2 \pi |\Gamma(s,x)|\right)^{d/2}}
        \exp\left(
            -\frac{1}{2} (z - s\meanq(x))^T \Gamma(s,x)^{-1} (z - s\meanq(x))
        \right) ,
    \end{align*}
    and note that if $\lambda_i$ are the eigenvalues of $\covq(x)$
    then $|\Gamma(s, x)| = \prod_i (\epsilon_r^2 + s \lambda_i)$,
    and
    $\partial_s |\Gamma(s, x)| = \sum_i \lambda_i |\Gamma(s, x)| / (\epsilon_r^2 + s \lambda_i)$.
    Therefore,
    \begin{align*}
        \partial_s p_{s, x}(z)
        &=
        \left(
            \meanq(x)^T \Gamma(s,x)^{-1} (z - s \meanq)
            +
            (z - s \meanq(x))^T
            \Gamma(s,x)^{-1} \covq(x) \Gamma(s,x)^{-1}
            (z - s \meanq(x))
        \right. \\ & \qquad \qquad \qquad \left. {}
            -
            \sum_i \frac{\lambda_i}{\epsilon_r^2 + s \lambda_i}
        \right)
        p_{s, x}(z) ,
    \end{align*}
    where $z^T$ is the transpose of $z$.
    This implies that
    \begin{align*}
        \frac{
        \theta
            \int_{\IR^d}
                ( p_{\epsilon_r^2}(y-w) - p_{\epsilon_r^2}(x-w) )
            q_\theta(x, dy)
        }{
            p_{\epsilon^2_\gamma}(x-w) .
        }
        =
        h(x-w) e^{k(x-w)},
    \end{align*}
    where $h(z)$ and $k(z)$ are quadratic polynomials in $z$
    whose coefficients depend on $s$ and $x$ but are uniformly bounded for $s \ge 1$
    and $x \in \IR^d$.
    Therefore, \eqref{eqn:goal1} holds for some finite $K$
    if and only if $k(z) < 0$ for all $z$ outside some bounded region.
    This is
    \begin{align*}
        k(z)
        =
        \frac{1}{2\epsilon_\gamma^2} \|z\|^2
        -\frac{1}{2} (z - s\meanq(x))^T \Gamma(s,x)^{-1} (z - s\meanq(x)) ,
    \end{align*}
    which is negative for all $z$ outside a bounded region.
    Since $\inf_z z^T \Gamma(s,x)^{-1} z / \|z\|^2 = 1 / (s \lambda(x) + \epsilon_r^2)$,
    where $\lambda(x) = \sup z^T \covq(x) z / \|z\|^2$ is the largest eigenvalue of $\covq(x)$,
    equation~\eqref{eqn:first_moment_rho} follows from the assumption
    that $\epsilon_r^2 + 2 \sup_x \lambda(x)/\theta < \epsilon_\gamma^2$.
    (Note that we do not yet need the factor of 2.)

    Next we prove equation~\eqref{eqn:second_moment_rho}, in a similar way.
    Again applying Fubini,
    \begin{align*}
        &
        \theta \int_{\IR^d}
                \left( \smooth{r} \eta(y) - \smooth{r} \eta(x) \right)^2
        q_\theta(x,dy)
        \\ &\qquad
        =
        \int_{\IR^d} \int_{\IR^d}
        \theta \int_{\IR^d}
            ( p_{\epsilon_r^2}(y - w) - p_{\epsilon_r^2}(x - w) )
            ( p_{\epsilon_r^2}(y - v) - p_{\epsilon_r^2}(x - v) )
        q_\theta(x, dy)
        \eta(dv) \eta(dw) ,
    \end{align*}
    and so as before, equation~\eqref{eqn:second_moment_rho} will follow if
    the integrand is bounded by $K p_\gamma(x-w) p_\gamma(x - v)$.
    % \begin{align*}
    %     \theta \int_{\IR^d}
    %         ( p_{\epsilon_r^2}(y - w) - p_{\epsilon_r^2}(x - w) )
    %         ( p_{\epsilon_r^2}(y - v) - p_{\epsilon_r^2}(x - v) )
    %     q_\theta(x, dy)
    %     \le K p_\gamma(x-w) p_\gamma(x - v) .
    % \end{align*}
    Now, let $Y_1$, $Y_2$, and $Z$ be independent $d$-dimensional Gaussians with mean zero,
    where $Y_1$ and $Y_2$ have covariance $\epsilon^2_r I$,
    and $Z$ has covariance $\covq(x)$.
    Write $p_{s,t,x}(\cdot, \cdot)$ for the joint density of
    $Y_1 + \sqrt{s} Z + s \meanq(x)$ and $Y_2 + \sqrt{t} Z + t \meanq(x)$.
    Then, observe that
    \begin{align*}
        &
        \theta \int_{\IR^d}
            ( p_{\epsilon_r^2}(y - w) - p_{\epsilon_r^2}(x - w) )
            ( p_{\epsilon_r^2}(y - v) - p_{\epsilon_r^2}(x - v) )
        q_\theta(x, dy)
        \\ &\qquad
        =
        \theta \left(
            p_{1/\theta, 1/\theta, x}(x-w, x-v)
            - p_{0, 1/\theta, x}(x-w, x-v)
        \right. \\ &\qquad \qquad \left. {}
            - p_{1/\theta, 0, x}(x-w, x-v)
            + p_{0, 0, x}(x-w, x-v)
        \right) 
        \\ &\qquad 
        =
            \partial_s p_{s, 1/\theta, x}(x-w, x-v)
            - \partial_t p_{0, t, x}(x-w, x-v) ,
    \end{align*}
    for some $0 \le s, t \le 1/\theta$.
    As before,
    \begin{align*}
        \frac{
            \theta \int_{\IR^d}
                ( p_{\epsilon_r^2}(y - w) - p_{\epsilon_r^2}(x - w) )
                ( p_{\epsilon_r^2}(y - v) - p_{\epsilon_r^2}(x - v) )
            q_\theta(x, dy)
        }{
            p_{\epsilon_\gamma^2}(x - w)
            p_{\epsilon_\gamma^2}(x - v)
        }
        =
        h(x-w, x-v) e^{ k(x-w, x-v) },
    \end{align*}
    where $h(z_1, z_2)$ is a polynomial with uniformly bounded coefficients and
    \begin{align*}
        k(z_1, z_2)
        =
        (\|z_1\|^2 + \|z_2\|^2)/(2 \epsilon_\gamma^2)
            - \frac{1}{2} [z_1, z_2[^T \Gamma(s, t, x)^{-1} [z_1, z_2] ,
    \end{align*}
    where $[z_1, z_2]$ is the $\IR^{2d}$ vector formed by concatenating $z_1$ and $z_2$,
    and $\Gamma(s, t, x)$ is the block matrix
    \begin{align*}
        \Gamma(s, t, x)
        =
        \left[
        \begin{array}{cc}
            \epsilon_r^2 I + s \covq(x) & \sqrt{st} \covq(x) \\
            \sqrt{st} \covq(x) & \epsilon_r^2 I + t \covq(x) \\
        \end{array}
        \right] .
    \end{align*}
    If $\covq(x) u = a u$ for some $a \in \IR$,
    then $[u \sqrt{s}, u \sqrt{t}]$ is an eigenvector of $\Gamma(s, t, x)$
    with eigenvalue $\epsilon^2_r + (s+t) a$,
    and $[u \sqrt{t}, - u \sqrt{s}]$ is an eigenvector of $\Gamma(s, t, x)$
    with eigenvalue 0.
    This implies the largest eigenvalue of $\Gamma(s, t, x)$
    is equal to $\epsilon^2_r + (s+t) \lambda(x)$,
    where $\lambda(x)$ is again the largest eigenvalue of $\covq(x)$.
    Therefore, if $s + t \le 2 / \theta$,
    \begin{align*}
        &
        (\|z_1\|^2 + \|z_2\|^2) / \epsilon^2_\gamma
            - [z_1, z_2[^T \Gamma(s, t, x)^{-1} [z_1, z_2]
        \\ &\qquad \le
        (\|z_1\|^2 + \|z_2\|^2) \left(
            \frac{1}{\epsilon^2_\gamma}
            - \frac{1}{\epsilon^2_r +  2\lambda(x)/\theta}
        \right) ,
    \end{align*}
    which is negative by assumption.
    Therefore, there is a $K$ such that
    \begin{align*}
        \frac{ \left|
            \partial_s p_{s, 1/\theta, x}(x-w, x-v)
            - \partial_t p_{0, t, x}(x-w, x-v)
        \right| }{
            p_{\epsilon^2_\gamma}(x - w)
            p_{\epsilon^2_\gamma}(x - v)
        }
        \le K
    \end{align*}
    for all $\theta > 1$ and all $x$, $v$, and $w \in \IR^d$,
    proving equation~\eqref{eqn:second_moment_rho} and hence the lemma.
